Ship-based lidar evaluation of Southern Ocean clouds in the storm-resolving general circulation model ICON, and the ERA5 and MERRA-2 reanalyses

### 2024-12-12 manuscript_agu.tex
Addressing comments

\begin{abstract}
Global storm-resolving models (GSRMs) are the next avenue of climate modeling. Among them is a 5-km version of the Icosahedral Nonhydrostatic Weather and Climate Model (ICON). The high resolution means that parameterizations of convection and clouds, including subgrid-scale clouds, are forgone, relying on explicit simulation, but still using microphysics and turbulence parametrization. Standard-resolution (10--100-km) models, which use subgrid-scale convection and cloud parametrization, have substantial cloud biases over the Southern Ocean (SO), affecting radiation and sea surface temperature. We evaluated SO clouds in ICON and the ERA5 and MERRA-2 reanalyses. The SO is dominated by low clouds, which cannot be observed accurately from space due to overlapping clouds, attenuation, and ground clutter. Instead, we analyzed about 2400 days of lidar observations from 31 voyages and a station during 2010--2021 using a ground-based lidar simulator. ICON and the reanalyses underestimate the total cloud fraction by about 10 and 20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at about 500 m, potentially explained by their lifting condensation levels being too high. The reanalyses strongly underestimate near-surface clouds or fog. MERRA-2 tends to underestimate cloud occurrence at all heights. In daily cloud cover, ICON and the reanalyses tend to be about 1 and 2 oktas clearer, respectively. Conditions of weak stability are associated with the largest errors in ICON and the reanalyses. Compared to radiosondes, potential temperature is accurate in the reanalyses to about 0.5 K, but ICON underestimates stability over the low-latitude SO and is too humid in the boundary layer. MERRA-2 is too humid at all heights. SO cloud biases remain a substantial issue in the GSRM but are an improvement over the lower-resolution reanalyses. Explicitly resolved convection and cloud processes (except microphysics) were not enough to address the model cloud biases.
\end{abstract}

@@

Global storm-resolving models are climate models with km-scale horizontal resolution, which are currently in development. Reanalyses are the best estimates of past meteorological conditions based on an underlying global model and observations. We evaluated clouds and thermodynamic profiles over the Southern Ocean in one such model, as well as two reanalyses, based on 2400 days of ship and station observations. Thanks to the high resolution, the model relies entirely on explicit simulation of clouds, instead of subgrid-scale parametrizations. For the evaluation, we used ceilometer and radiosonde observations and a lidar simulator, which enables a fair comparison with the model and reanalyses. We subsetted our results by cyclonic activity and stability. We found that the model and reanalyses underestimate a lidar-derived cloud fraction, and the reanalyses do so more strongly. Fog or near-surface clouds are especially underestimated in the reanalyses. However, the model and one of the reanalyses also tend to overestimate the peak of cloud occurrence at 500 m above the ground, and it tends to be higher. This is linked to thermodynamic profiles, which show a higher lifting condensation level. Southern Ocean biases are still an important problem in the model, but an improvement over the reanalyses is notable.

@@

The main aim of this study is to evaluate the GSRM version of ICON. ICON is developed and maintained jointly by Deutscher Wetterdienst, Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Previous studies have identified substantial large-scale biases in climate model clouds over the SO, affecting sea surface temperature and the Earth’s albedo. Our aim is to quantify how well the GSRM ICON is simulating clouds in this region, particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on CALIPSO (neither of which are now operational) are affected by the abovementioned issues, resulting in a strong underestimation of cloud occurrence below 2 km relative to ground-based lidar observations \cite{mcerlich2021}. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parametrized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the RV \emph{Polarstern} and other voyages and stations as a reference for the model evaluation.

@@

\subsection{Voyage and station data}

@@

The Modern-Era Retrospective analysis for Research and Applications, Version 2 (MERRA-2) is a reanalysis produced by the Global Modeling and Assimilation Office at the NASA Goddard Space Flight Center \cite{gelaro2017}. It uses version 5.12.4 of the Goddard Earth Observing System (GEOS) atmospheric model \cite{rienecker2008,molod2015}. Non-convective clouds (condensation, autoconversion, and evaporation) are parametrized using a prognostic scheme \cite{bacmeister2006}, and sub-grid cloud fraction is determined using total water distribution and a critical relative humidity threshold. The reanalysis output analyzed here is available at a spatial resolution of 0.5° of latitude and 0.625° of longitude, which is about 56 km in the North--South direction and 35 km in the East--West direction at 60°S. The number of vertical model levels is 72. Here, we use the following products: 1-hourly instantaneous 2D single-level diagnostics (M2I1NXASM) for 2-m temperature and humidity; 3-hourly instantaneous 3D assimilated meteorological fields (M2I3NVASM) for cloud quantities, pressure, and temperature; 1-hourly average 2D surface flux diagnostics (M2T1NXFLX) for precipitation; and 1-hourly average 2D radiation diagnostics (M2T1NXRAD) for radiation quantities \cite{merra2}.

@@

\subsection{Precipitation identification using machine learning}
\label{sec:ann}

@@

\subsection{Partitioning by cyclonic activity and stability}
\label{sec:cyclone-stability}

@@

\subsection{Cyclonic activity and stability}

@@

\subsection{Cloud occurrence by height}
\label{sec:cloud-occurrence}

@@

\subsection{Cloud cover}
\label{sec:cloud-cover}

@@

\subsection{Thermodynamic profiles}

@@

\section{Limitations of this study}
\label{sec:limitations}

@@

\section{Discussion and conclusions}

@@

The radiosonde observations indicate that the LCL is too high in ICON and reanalyses, which is probably responsible for the higher peak of clouds in the models and the lack of near-surface clouds or fog. The radiosonde comparison, however, does not seem to explain cloud biases at higher altitudes. MERRA-2 is too moist at all heights. ICON also exhibits smaller internal variability than the radiosonde observations. Overall, the radiosonde comparison only partially explains the identified cloud biases, and other physical causes are likely contributing. This warrants further investigation, especially of ocean--atmosphere fluxes, shallow convection, and boundary layer turbulence. The lack of parametrized subgrid-scale convection in ICON could be a substantial issue even at the 5-km resolution.

@@

The results imply that SO cloud biases are still a substantial issue even in the km-scale resolution ICON model, even though an improvement over the lower-resolution reanalyses is notable. More effort is needed to improve the model cloud simulations in this understudied region. However, this analysis suggests that the transition from models with parametrized convection and clouds to storm-resolving models might not solve these biases without additional effort. Evaluation of ocean--atmosphere heat, moisture, and momentum fluxes against in-situ observations over the SO and comparison of GSRM simulations against large-eddy simulations are two potential avenues for future research that could elucidate the physical mechanisms behind the biases, in addition to the more common efforts in SO cloud microphysics and precipitation evaluation.

>>>>

\begin{abstract}
Global storm-resolving models (GSRMs) are the upcoming global climate models. One of them is a 5-km Icosahedral Nonhydrostatic Weather and Climate Model (ICON). The high resolution means that parameterizations of convection and clouds, including subgrid-scale clouds, are omitted, relying on explicit simulation but still utilizing microphysics and turbulence parameterizations. Standard-resolution (10--100 km) models, which use convection and cloud parameterizations, have substantial cloud biases over the Southern Ocean (SO), affecting radiation and sea surface temperature. The SO is dominated by low clouds, which cannot be observed accurately from space due to overlapping clouds, attenuation, and ground clutter. We evaluated SO clouds in ICON and the ERA5 and MERRA-2 reanalyses using about 2400 days of lidar observations and 2300 radiosonde profiles from 31 voyages and a station during 2010--2021, compared with the models using a ground-based lidar simulator. We found that ICON and the reanalyses underestimate the total cloud fraction by about 10 and 20\%, respectively. ICON and ERA5 overestimate the cloud occurrence peak at about 500 m, potentially explained by their lifting condensation levels being too high. The reanalyses strongly underestimate fog or near-surface clouds and MERRA-2 underestimates cloud occurrence at almost all heights. Outgoing shortwave radiation is overestimated in the reanalyses, implying a ``too few, too bright'' cloud problem. Thermodynamic conditions are relatively well represented, but ICON is less stable than observations and MERRA-2 is too humid. SO cloud biases are a substantial issue in the GSRM, but it matches the observations better than the lower-resolution reanalyses.
\end{abstract}

@@

Global storm-resolving models are climate models with km-scale horizontal resolution, which are currently in development. Reanalyses are the best estimates of past meteorological conditions based on an underlying global model and observations. We evaluated clouds and thermodynamic profiles over the Southern Ocean in one such model, as well as two reanalyses, based on 2400 days of ship and station observations. Thanks to the high resolution, the model relies entirely on explicit simulation of clouds, instead of subgrid-scale parameterizations. For the evaluation, we used ceilometer and radiosonde observations and a lidar simulator, which enables a fair comparison with the model and reanalyses. We subsetted our results by cyclonic activity and stability. We found that the model and reanalyses underestimate a lidar-derived cloud fraction, and the reanalyses do so more strongly. Fog or near-surface clouds are especially underestimated in the reanalyses. However, the model and one of the reanalyses also tend to overestimate the peak of cloud occurrence at 500 m above the ground, and it tends to be higher. This is linked to thermodynamic profiles, which show a higher lifting condensation level. Southern Ocean biases are still an important problem in the model, but an improvement over the reanalyses is notable.

@@

The main aim of this study is to evaluate the GSRM version of ICON. ICON is developed and maintained jointly by Deutscher Wetterdienst, Max-Planck-Institute for Meteorology, Deutsches Klimarechenzentrum (DKRZ), Karlsruhe Institute of Technology, and the Center for Climate Systems Modeling. Previous studies have identified substantial large-scale biases in climate model clouds over the SO, affecting sea surface temperature and the Earth’s albedo. Our aim is to quantify how well the GSRM ICON is simulating clouds in this region, particularly in light of the fact that subgrid-scale clouds and convection are not parameterized in this model. This region is mostly dominated by boundary layer clouds generated by shallow convection, and these are problematic to observe by spaceborne lidars and radars, which are affected by attenuation by overlapping and thick clouds \cite{mace2009,medeiros2010} and ground clutter \cite{marchand2008}, respectively. Specifically, the radar on CloudSat and lidar on CALIPSO (neither of which are now operational) are affected by the abovementioned issues, resulting in a strong underestimation of cloud occurrence below 2 km relative to ground-based lidar observations \cite{mcerlich2021}. We hypothesize that this, in turn, can lead to systematic biases in low clouds in climate models, which are frequently evaluated against CloudSat--CALIPSO products. Reanalyses can also suffer from cloud biases, as these are usually parameterized in their atmospheric component and also in regions where input observations are sparse. This makes them a problematic reference for clouds over the SO, and any biases relative to a reanalysis should be interpreted with caution. Instead, we chose to use a large set of ship-based observations conducted with ceilometers and lidars on board the RV \emph{Polarstern} and other voyages and stations as a reference for the model evaluation.

@@

\subsection{Voyage and Station Data}

@@

The Modern-Era Retrospective analysis for Research and Applications, Version 2 (MERRA-2) is a reanalysis produced by the Global Modeling and Assimilation Office at the NASA Goddard Space Flight Center \cite{gelaro2017}. It uses version 5.12.4 of the Goddard Earth Observing System (GEOS) atmospheric model \cite{rienecker2008,molod2015}. Non-convective clouds (condensation, autoconversion, and evaporation) are parameterized using a prognostic scheme \cite{bacmeister2006}, and sub-grid cloud fraction is determined using total water distribution and a critical relative humidity threshold. The reanalysis output analyzed here is available at a spatial resolution of 0.5° of latitude and 0.625° of longitude, which is about 56 km in the North--South direction and 35 km in the East--West direction at 60°S. The number of vertical model levels is 72. Here, we use the following products: 1-hourly instantaneous 2D single-level diagnostics (M2I1NXASM) for 2-m temperature and humidity; 3-hourly instantaneous 3D assimilated meteorological fields (M2I3NVASM) for cloud quantities, pressure, and temperature; 1-hourly average 2D surface flux diagnostics (M2T1NXFLX) for precipitation; and 1-hourly average 2D radiation diagnostics (M2T1NXRAD) for radiation quantities \cite{merra2}.

@@

\subsection{Precipitation Identification Using Machine Learning}
\label{sec:ann}

@@

\subsection{Partitioning by Cyclonic Activity and Stability}
\label{sec:cyclone-stability}

@@

\subsection{Cyclonic Activity and Stability}

@@

\subsection{Cloud Occurrence by Height}
\label{sec:cloud-occurrence}

@@

\subsection{Top of Atmosphere Radiation}

In Fig.\ \ref{fig:cloud-occurrence}, we also show outgoing shortwave and longwave top of atmosphere radiation. In observations, these come from daily mean CERES measurements averaged over the voyage tracks or a station location, whereas in the models they come from daily means of TOA radiation in the model output averaged over the same location and time periods. In the free-running ICON model, the time period is mapped onto the available years, as explained in Section\ \ref{sec:icon}.

In the general case (Fig.\ \ref{fig:cloud-occurrence}a), ICON underestimates the outgoing SW radiation by 26 Wm$^{-2}$ and the reanalyses MERRA-2 and ERA5 are much closer, but overestimate it by 6 and 14 Wm$^{-2}$, respectively. While in ICON, this is in line with the underestimated total cloud fraction by 10\%, in the reanalyses this is the opposite result as expected from the underestimated total cloud fraction by about 20\%. The likely explanation is an overestimated cloud albedo, compensating for the lack of cloud area.

In contrast, LW radiation if affected by much smaller biases than SW radiation, which is expected due to the prevailing low-level clouds having similar temperature as the surface. In ICON, the outgoing LW radiation is overestimates by 8\%, which could be caused by understimated total cloud fraction exposing a larger area sea surface are to cooling to space, which is typically warmer than the atmospheric temperature at 0--2 km, where most of the clouds are located. In the MERRA-2 and ERA5 reanalyses, the LW biases are also slighly positive by 4 and 7 Wm$^{-2}$. This is again in line with the underestimated total cloud fraction by about 20\%. However, if the clouds are too thick, as expected from the SW results, this might also provide a compensating effect where too small cloud area is counteracted by greater thermal emissivity, thus reducing the ougoing LW radiation more relative to thinner clouds. For thin clouds, the outgoing TOA LW radiation originates both from the warmer surface (partly blocked by the clouds) and the clouds, whereas for thick clouds, the outgoing TOA LW radiation originates mostly from the colder-than-surface clouds.

In all of the subsets (Fig.\ \ref{fig:cloud-occurrence}b--g), the same type of bias is present as in the general case -- the outgoing SW radiation is underestimated in ICON and overestimated in MERRA-2 and ERA5, and the outgoing LW radiation is overestimated in all of the models. Even though the total cloud fraction is lower by 7\% over the high-latitude SO than the low-latitude SO, the outgoing SW radiation is much greater by 41 Wm$^{-2}$, implying a much greater cloud albedo over the high-latitude SO. ICON has the same contrast between these two regions in the total cloud fraction and SW radiation, but the outgoing SW radiation difference between the regions is much smaller (16 Wm$^{-2}$). The reanalyses do not show this type of contrast between the regions. The physical reason for this might be that the prevalence of fog or low-level clouds over the low-latitude SO and their relative lack over the high-latitude SO in obserations is not reproduced in the models (Fig.\ \ref{fig:cloud-occurrence}b--c).

\subsection{Cloud Cover}
\label{sec:cloud-cover}

@@

\subsection{Thermodynamic Profiles}

@@

\section{Limitations of this Study}
\label{sec:limitations}

@@

\section{Discussion and Conclusions}

@@

The radiosonde observations indicate that the LCL is too high in ICON and reanalyses, which is probably responsible for the higher peak of clouds in the models and the lack of near-surface clouds or fog. The radiosonde comparison, however, does not seem to explain cloud biases at higher altitudes. MERRA-2 is too moist at all heights. ICON also exhibits smaller internal variability than the radiosonde observations. Overall, the radiosonde comparison only partially explains the identified cloud biases, and other physical causes are likely contributing. This warrants further investigation, especially of ocean--atmosphere fluxes, shallow convection, and boundary layer turbulence. The lack of parameterized subgrid-scale convection in ICON could be a substantial issue even at the 5-km resolution.

@@

The results imply that SO cloud biases are still a substantial issue even in the km-scale resolution ICON model, even though an improvement over the lower-resolution reanalyses is notable. More effort is needed to improve the model cloud simulations in this understudied region. However, this analysis suggests that the transition from models with parameterized convection and clouds to storm-resolving models might not solve these biases without additional effort. Evaluation of ocean--atmosphere heat, moisture, and momentum fluxes against in-situ observations over the SO and comparison of GSRM simulations against large-eddy simulations are two potential avenues for future research that could elucidate the physical mechanisms behind the biases, in addition to the more common efforts in SO cloud microphysics and precipitation evaluation.
